Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Flows

### 2016-07-03 main.tex
finished transformation matrix
More or less finished first draft of methods section Starting results section

@@

\usepackage[affil-it]{authblk}
\usepackage{lineno}
\usepackage{soul}
\usepackage{color}
\usepackage[utf8]{inputenc}
\usepackage[square,numbers,sort&compress]{natbib}

\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{mathabx}
\usepackage{bm}

\newcommand{\pos}{\mathbf{x}}
\newcommand{\pvel}{\boldsymbol{\xi}}
\newcommand{\mvel}{\mathbf{u}}
\newcommand{\cop}{\boldsymbol{\Omega}}
\newcommand{\grad}{\boldsymbol{\nabla}}
\newcommand{\mmom}{\mathbf{j}}

\title{Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Flow}
%\title{A Comparison of Lattice Boltzmann Method Collision Operators for Accuracy, Stability and Efficiency in Simulating Non-Newtonian Flows}

@@

\linenumbers

@@

The dynamic viscosity of a fluid is a measure of its resistance to shear deformation.
For many fluids, at a constant temperature, the dynamic viscosity can be considered as constant.
These fluids are known as Newtonian fluids.
Non-Newtonian fluids have an apparent dynamic viscosity, or apparent resistance to shear stress, that is variable--even at a constant temperature--and is often a function of strain-rate.
There are a number of fluids in science and engineering applications that can be classified as non-Newtonian; pastes, slurries, molten plastics, polymer solutions, dyes, varnishes, suspensions, and some biomedical liquids such as blood all behave in a non-Newtonian manner \cite{bohme1987non}.
Of all of the different non-Newtonian behaviors that exist, there are two models under which much of the behaviors may be idealized:
\begin{enumerate}
	\item \emph{Yield stress fluids}. Yield stress fluids, also known as Bingham plastics, do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
	Some examples of yield stress fluids are drilling muds, fresh cement slurries and concretes, and melted metal alloys \cite{ginzburg2002free}.
	Understanding the flow of cement slurries and concretes is important because they are used in the cementing of oil and gas wellbores, and in construction.
	If the cement job for an oil and gas wellbore is not completed successfully then gases and fluids can leak through the wellbore to contaminate ground water or the atmosphere.
	The goal of understanding cement slurry and concrete flows during construction is to understand how construction forms can be completely filled as voids can lead to structural deficiencies~\cite{definitely cite someone}.
	Yield stress flow is relevant in many other disciplines and applications because of the many substances that exhibit yield stress behavior, e.g. pastes, paints, muds, molten plastics and metals, and blood \cite{wang2011lattice}. %MG: citation is for blood as non-Newtonian
	As a consequence, simulating yield stress fluids can help engineers develop better pastes and paints, understand the flow and deformation of mud and clay in geotechnical engineering, understand blood circulation, and better manufacture metal and plastic parts.

	\item \emph{Power-law fluids}. Power-law behavior is more commonly known as \emph{shear-thinning}--when the apparent viscosity decreases with increasing strain-rate--or \emph{shear-thickening}--when the apparent viscosity increases with increasing strain-rate. Shear-thinning fluids are also known as \emph{pseudoplastics}, and some examples include polymer mixes and molten plastics.
	Shear-thickening fluids are also known as \emph{dilatants}, and some examples include quicksand, a cornstarch and water mixture, and a silica and polyethylene glycol mixture.
\end{enumerate} %MG: should HB, K-L, Casson, Carreau, and other models be mentioned?

@@

\subsection{Lattice Boltzmann Method for non-Newtonian Flow} %MG: should this be broken down into subsubsections? Maybe like 'Motivation', 'Stability', 'Recent Developments to Address Stability'

Analytical solutions rarely exists for even the simplest non-Newtonian flows because of the complexity that a non-linear constitutive relationship entails.
It is generally more practical to approximate non-Newtonian flows using numerical methods.
However, the non-linear constitutive equation--typically of the form $\tau = \mu_{app}(\dot{\gamma}) \dot{\gamma}$, where the apparent viscosity, $\mu_{app}$ is a function of the strain rate--results in certain challenges for numerical methods as well.
%Constitutive form for generalized Newtonian fluids, $\tau = \mu(\dot{\gamma}) \dot{\gamma}$, where the apparent viscosity, $\mu$ is a function of the strain rate
Determining the apparent viscosity and shear rate of a flow will often require an iterative solution, a general Picard-type algorithm is:

@@

Stability concerns are just as prevalent, if not more prevalent, in simulating non-Newtonian flows because the strong nonlinear relationship between shear stress and strain-rate can lead to high non-equilibrium fluctuations. %MG: be more precise than 'fluctuations'?

@@

A potential issue with the stability of the variable relaxation time approach is that as the relaxation time, $\tau$, approaches $1/2$ the viscosity approaches zero; and alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method also degrades~\cite{latt2007hydrodynamic}.

@@

Much work has been done (not specific to non-Newtonian flows) to investigate and develop ways to enhance the stability of lattice Boltzmann methods.
\citet{sterling1993stability,sterling1996stability} investigated the stability of LBM with the linear Bhatnagar--Gross--Krook (BGK) collision operator on 7-velocity and 9-velocity 2D lattices and a 15-velocity cubic lattice.
The investigation was performed by linearizing all nonlinear terms about global equilibrium values and applying a von Neumann stability analysis on the resulting linearized equations.
The spatial dependence of stability was analyzed by taking the Fourier transform of the linearized equations.
\citeauthor{sterling1993stability} concluded that the linear stability of LBM models depends on the mass distribution parameters, the mean velocity, the relaxation time, and the wave number. %MG: this probably requires some clarification and a more concrete tieback to the highlevel goals of this paper
%(Some misc notes from this study that probably don't fit here but are important to this work: ``... a Courant stability condition is superceded by a more stringent stability condition on the speeds. Another stability boundary common for finite-difference methods requires that the viscous diffusion speed be less than the lattice spacing divided by the time step. This boundary is not observed for LB methods because as the viscosity increases, errors of the scheme increase due to the presence of large nonequilibirum populations, but stability is still maintained''??).
~\citet{worthing1997stability} extended the stability analysis performed by \citeauthor{sterling1993stability}, and instead of linearizing the LBM equations by expanding about global equilibrium values, \citeauthor{worthing1997stability} \ul{does something else}.
Their analysis discover both physical and nonphysical instabilities, and that setting the BGK relaxation time, $\tau = 1$ provides the optimal accuracy in time. %MG: is this actually relevant though?
\ul{Expand a little more about what they did and its significance here.}
%MG: all this stability talk is nice, but I need to figure out how to tie it back to non-Newtonian flows sooner rather than later. The problem is a lot of these studies focus on instability due to high Reynolds number flows, not really our problem. We're more concerned with instability due to higher relaxation rates and apparent viscosities. And how does MRT address this instability?

@@

To improve upon the stability of LBM models, \citet{d1994generalized} developed a multiple-relaxation-time (MRT) collision operator, that takes place in moment space and allows each moment to relax at a different rate.
\citet{lallemand2000theory} used von Neumann stability analysis to investigate the dispersion, dissipation, isotropy, Galilean invariance, and stability of the newly constructed LBM-MRT model.
They concluded LBM with the MRT collision operator was more stable, but with increased computational expense, than what is still often used, the BGK collision operator.
Note that although this increased computational expense was decided to be not too significant for Newtonian flows ($\approx$ 10-20\% \cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flows because an iterative solution for the constitutive equation can require that certain aspects of the collision operator be performed at each iteration (the strain rate tensor is determined from the non-equilibrium distribution function which must be mapped into momentum space when using the MRT collision operator).
Both the BGK~\cite{wang2011lattice,ashrafizaadeh2009comparison,wang2015localized,tang2011bingham} and MRT~\cite{chen2014simulations,fallah2012multiple,chai2011multiple} collision operators have been used successfully to simulate non-Newtonian flows.
\citeauthor{chen2014simulations} concluded that the MRT collision operator was more stable for Bingham plastic flow and allowed the use of a more accurate approximation to the Bingham plastic constitutive relationship.

{\color{red} 
ELBM and entropic filtering stuff goes here.
The price of entropic lattice Boltzmann models for fluid flow is living with a bounded transport coefficient.
Entropically stabilized collision operators ensure that entropy always increases, i.e. that the second law of thermodynamics is never violated.
For BGK, this means that the relaxation time is bounded from below in certain instances~\cite{boghosian2001entropic}.
``it may be argued that we are relinquishing some control over the transport coefficients'' and ``of course, making $\tau$ a nontrivial function of the incoming state will impact the hydrodynamic equations derived from the model''~\cite{boghosian2001entropic}.
``stability is guaranteed''?
Define Lyapunov functional or Lyapunov stability condition.
}

@@

\section{Lattice Boltzmann Method}

\subsection{Overview}

Some overview of Lattice Boltzmann Method stuff~\cite{grasinger2015simulation}.

@@

\subsection{The Boltzmann Equation}
The Boltzmann equation (BE) can be thought of as a conservation of particle distributions.
The BE is given as $\frac{\partial f}{\partial t} + \pvel \cdot \grad f = \cop$, where $f = f(\pos, \pvel, t)$ is the particle velocity distribution function, $\pos$ is the spatial position vector, $\pvel$ is the particle velocity, $\mvel$ is the macroscopic velocity, and $\cop$ is the collision operator.
The lattice part of LBM refers to the way in which the BE is discretized.
The lattice discretizes the spatial domain with nodes that are connected to their neighbors through discrete lattice velocity vectors.
The velocity vectors act as pathways for particle distributions to travel along.
Each time step in LBM consists of two distinct actions:

@@

The D2Q9 lattice was used in the current work (shown in Figure \ref{fig:d2q9}), which is commonly used for two-dimensional, incompressible flow simulations~\cite{Suc01}. % MG: do we even need this citation?
The lattice is two-dimensional with nine discrete velocities at each node.
There is a stationary particle, there are four discrete velocities of magnitude 1, $(\pm1,0)$, $(0,\pm1)$, and there are four discrete velocities of magnitude $\sqrt{2}$, $(\pm1, \pm1)$.

@@

The discretized version of the Boltzmann equation, or the lattice Boltzmann equation (LBE), is given as: $f_i(\pos + \pvel_i, t + \Delta t) = f_i(\pos, t) + \cop_i(f)$, where, for D2Q9, $i = 0, 1, ..., 8$.

@@

No slip, or zero velocity, which is commonly imposed at walls in a domain, is accomplished by simulating the particle distributions as "bouncing back" at the walls in the opposite direction from which they stream.
For example, for a particle distribution streaming in the direction of a south wall, $f_2 = f_4$, $f_5 = f_7$, and $f_6 = f_8$.
For velocity or pressure boundary conditions, the method proposed by \citet{zou1997pressure} can be used.
The particle distributions that are missing after the streaming step are solved for by assuming a bounceback of the non-equilibrium distribution in the direction normal to the boundary; e.g., for a south inlet or outlet, $f_2 - f_2^{eq} = f_4 - f_4^{eq}$.

@@

\subsection{Collision Operator} %MG: if time allows, checkout differential lattice Bolzmann equation, the interpolation-supplemented lattice Boltzmann method, and the Taylor series expansion- and least square-based lattice Boltzmann method; Niu et. al. 2004

The collision operator, in the case of the continuous BE, attempts to describe the change in particle momentums and trajectories due to pairwise particle collisions (based on their respective momentums and trajectories just prior to collision) \cite{Cer90}.
In LBM, the collision operator causes particle distributions to relax toward a local equilibrium. This equilibrium is determined by the macroscopic physical behavior of interest.
In the case of incompressible flow, the equilibrium particle distribution, $f_i^{eq}$, is given by:

\begin{equation}
f_i^{eq} = w_i \rho \left[1 + \frac{\pvel_i \cdot \mvel}{c_s^2} + \frac{(\pvel_i \cdot \mvel)^2}{2c_s^2} - \frac{\mvel^2}{2c_s^2} \right]
\end{equation}

\noindent where $w_i$ is the weight in the $i^{th}$ direction.

\begin{equation*}
w_i = \begin{cases}
    \frac{4}{9}, & i = 0 \\
    \frac{1}{9}, & i = 1, 2, 3, 4 \\
    \frac{1}{36}, & i = 5, 6, 7, 8
\end{cases}
\end{equation*}

@@

\begin{equation}
\Omega_i = -\frac{1}{\tau} (f_i(\pos, t) - f_i^{eq}),
\end{equation}

\noindent where $\tau$ is the relaxation time \cite{Bha54}.
The relaxation time can be related to macroscopic constitutive properties through Chapman-Enskog multiscale analysis.
For incompressible Newtonian flow, the relaxation time is related to the kinematic viscosity by $\nu = c_s^2(\tau - \frac{1}{2})$.

\subsubsection{Entropic Collisions and Entropy Balance}

>>>>

% MG: replace \ \cite with ~\cite to ensure citations stay on same line
% MG: need an overall consistency in 'flucuation' vs. 'oscillation'; use same word if you mean same thing.
% MG: same note as above for 'local equilibrium' vs. 'quasiequilibrium'
% MG: consider using collision frequency, \omega, instead of relaxation time, \tau, so that the relaxation time is not confused with shear stress

@@

\usepackage[affil-it]{authblk}
\usepackage{lineno}
\usepackage{soul}
\usepackage{color}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[square,numbers,sort&compress]{natbib}

\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{mathabx}
\usepackage{bm}
\usepackage[plain]{fancyref}

\newcommand{\pos}{\mathbf{x}}
\newcommand{\pvel}{\boldsymbol{\xi}}
\newcommand{\mvel}{\mathbf{u}}
\newcommand{\colop}{\boldsymbol{\Omega}}
\newcommand{\grad}{\boldsymbol{\nabla}}
\newcommand{\mmom}{\mathbf{j}}
\newcommand{\transM}{\mathbf{M}}
\newcommand{\relaxM}{\mathbf{S}}

\newcommand{\expnumber}[2]{{#1}\mathrm{e}{#2}}

\title{Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Flows}
%\title{A Comparison of Lattice Boltzmann Method Collision Operators for Accuracy, Stability and Efficiency in Simulating Non-Newtonian Flows}

@@

@@

The dynamic viscosity of a fluid is a measure of its resistance to shear deformation.
For many fluids, at a constant temperature, the dynamic viscosity can be considered as constant.
These fluids are known as Newtonian fluids.
Non-Newtonian fluids have an apparent dynamic viscosity, or apparent resistance to shear stress, that is variable--even at a constant temperature--and is often a function of strain-rate.
There are a number of fluids in science and engineering applications that can be classified as non-Newtonian; pastes, slurries, molten plastics, polymer solutions, dyes, varnishes, suspensions, and some biomedical liquids such as blood all behave in a non-Newtonian manner \cite{bohme1987non}.
Of all of the different non-Newtonian behaviors that exist, there are two models under which much of the behaviors may be idealized:
\begin{enumerate}
	\item \emph{Yield stress fluids}. Yield stress fluids, also known as Bingham plastics, do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
	Some examples of yield stress fluids are drilling muds, fresh cement slurries and concretes, and melted metal alloys \cite{ginzburg2002free}.
	Understanding the flow of cement slurries and concretes is important because they are used in the cementing of oil and gas wellbores, and in construction.
	If the cement job for an oil and gas wellbore is not completed successfully then gases and fluids can leak through the wellbore to contaminate ground water or the atmosphere~\cite{grasinger2015simulation}. %MG: is it tacky to cite myself here?
	The goal of understanding cement slurry and concrete flows during construction is to understand how construction forms can be completely filled as voids can lead to structural deficiencies~\cite{definitely cite someone}.
	Yield stress flow is relevant in many other disciplines and applications because of the many substances that exhibit yield stress behavior, e.g. pastes, paints, muds, molten plastics and metals, and blood \cite{wang2011lattice}. %MG: citation is for blood as non-Newtonian
	As a consequence, simulating yield stress fluids can help engineers develop better pastes and paints, understand the flow and deformation of mud and clay in geotechnical engineering, understand blood circulation, and better manufacture metal and plastic parts.

	\item \emph{Power-law fluids}. Power-law behavior is more commonly known as \emph{shear-thinning}--when the apparent viscosity decreases with increasing strain-rate--or \emph{shear-thickening}--when the apparent viscosity increases with increasing strain-rate. Shear-thinning fluids are also known as \emph{pseudoplastics}, and some examples include polymer mixes and molten plastics.
	Shear-thickening fluids are also known as \emph{dilatants}, and some examples include quicksand, a cornstarch and water mixture, and a silica and polyethylene glycol mixture.
\end{enumerate}

@@

\subsection{Lattice Boltzmann Method for non-Newtonian Flow} \label{sec:lbm-for-nnf} %MG: should this be broken down into subsubsections? Maybe like 'Motivation', 'Stability', 'Recent Developments to Address Stability'

Analytical solutions rarely exists for even the simplest non-Newtonian flows because of the complexity that a nonlinear constitutive relationship entails.
It is generally more practical to approximate non-Newtonian flows using numerical methods.
However, the nonlinear constitutive equation--typically of the form \begin{equation} \label{eq:non-newtonian}
\tau = \mu_{app}(\dot{\gamma}) \dot{\gamma},
\end{equation}\noindent where the apparent viscosity, $\mu_{app}$ is a function of the strain rate--results in certain challenges for numerical methods as well.
%Constitutive form for generalized Newtonian fluids, $\tau = \mu(\dot{\gamma}) \dot{\gamma}$, where the apparent viscosity, $\mu$ is a function of the strain rate
Determining the apparent viscosity and shear rate of a flow will often require an iterative solution, a general Picard-type algorithm is:

@@

Stability concerns are just as prevalent, if not more prevalent, in simulating non-Newtonian flows because the strong nonlinear relationship between shear stress and strain-rate can lead to high nonequilibrium fluctuations. %MG: be more precise than 'fluctuations'?

@@

A potential issue with the stability of the variable relaxation time approach is that as the collision frequency, $\omega$, approaches $2$ the viscosity approaches zero; and alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method also degrades~\cite{latt2007hydrodynamic}.

@@

Much work has been done (not specific to non-Newtonian flows) to investigate and develop ways to enhance the stability of lattice Boltzmann methods.
\citet{sterling1993stability,sterling1996stability} investigated the stability of LBM with the linear Bhatnagar--Gross--Krook (BGK) collision operator on 7-velocity and 9-velocity 2D lattices and a 15-velocity cubic lattice.
The investigation was performed by linearizing all nonlinear terms about global equilibrium values and applying a von Neumann stability analysis on the resulting linearized equations.
The spatial dependence of stability was analyzed by taking the Fourier transform of the linearized equations.
\citeauthor{sterling1993stability} concluded that the linear stability of LBM models depends on the mass distribution parameters, the mean velocity, the relaxation time, and the wave number. %MG: this probably requires some clarification and a more concrete tieback to the highlevel goals of this paper
%(Some misc notes from this study that probably don't fit here but are important to this work: ``... a Courant stability condition is superceded by a more stringent stability condition on the speeds. Another stability boundary common for finite-difference methods requires that the viscous diffusion speed be less than the lattice spacing divided by the time step. This boundary is not observed for LB methods because as the viscosity increases, errors of the scheme increase due to the presence of large nonequilibirum populations, but stability is still maintained''??).
~\citet{worthing1997stability} extended the stability analysis performed by \citeauthor{sterling1993stability}, and instead of linearizing the LBM equations by expanding about global equilibrium values, \citeauthor{worthing1997stability} \ul{does something else}.
Their analysis discover both physical and nonphysical instabilities, and that setting the BGK collision frequency, $\omega = 1$ provides the optimal accuracy in time. %MG: is this actually relevant though?
\ul{Expand a little more about what they did and its significance here.}
%MG: all this stability talk is nice, but I need to figure out how to tie it back to non-Newtonian flows sooner rather than later. The problem is a lot of these studies focus on instability due to high Reynolds number flows, not really our problem. We're more concerned with instability due to lower relaxation rates and apparent viscosities. And how does MRT address this instability?

@@

To improve upon the stability of LBM models, \citet{d1994generalized} developed a multiple-relaxation-time (MRT) collision operator, that takes place in moment space and allows each moment to relax at a different rate.
\citet{lallemand2000theory} used von Neumann stability analysis to investigate the dispersion, dissipation, isotropy, Galilean invariance, and stability of the newly constructed LBM-MRT model.
They concluded LBM with the MRT collision operator was more stable, but with increased computational expense, than what is still often used, the BGK collision operator.
Note that although this increased computational expense was decided to be not too significant for Newtonian flows ($\approx$ 10-20\% \cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flows because an iterative solution for the constitutive equation can require that certain aspects of the collision operator be performed at each iteration (the strain rate tensor is determined from the nonequilibrium distribution function which must be mapped into momentum space when using the MRT collision operator).
Both the BGK~\cite{wang2011lattice,ashrafizaadeh2009comparison,wang2015localized,tang2011bingham} and MRT~\cite{chen2014simulations,fallah2012multiple,chai2011multiple} collision operators have been used successfully to simulate non-Newtonian flows.
\citeauthor{chen2014simulations} concluded that the MRT collision operator was more stable for Bingham plastic flow and allowed the use of a more accurate approximation to the Bingham plastic constitutive relationship.

{\color{red} 
ELBM and entropic filtering stuff goes here.
The price of entropic lattice Boltzmann models for fluid flow is living with a bounded transport coefficient.
Entropically stabilized collision operators ensure that entropy always increases, i.e. that the second law of thermodynamics is never violated.
For BGK, this means that the relaxation time is bounded from below in certain instances~\cite{boghosian2001entropic}.
``it may be argued that we are relinquishing some control over the transport coefficients'' and ``of course, making $\omega$ a nontrivial function of the incoming state will impact the hydrodynamic equations derived from the model''~\cite{boghosian2001entropic}.
``stability is guaranteed''?
Define Lyapunov functional or Lyapunov stability condition.
}

@@

\section{Numerical Methods}

\subsection{Lattice Boltzmann Method: Overview}

@@

\subsection{The Boltzmann Equation}
The Boltzmann equation (BE) can be thought of as a conservation of particle distributions.
The BE is given as $\frac{\partial f}{\partial t} + \pvel \cdot \grad f = \colop$, where $f = f(\pos, \pvel, t)$ is the particle velocity distribution function, $\pos$ is the spatial position vector, $\pvel$ is the particle velocity, $\mvel$ is the macroscopic velocity, and $\colop$ is the collision operator.
The lattice part of LBM refers to the way in which the BE is discretized.
The lattice discretizes the spatial domain with nodes that are connected to their neighbors through discrete lattice velocity vectors.
The velocity vectors act as pathways for particle distributions to travel along.
Each time step in LBM consists of two distinct actions:

@@

The D2Q9 lattice was used in the current work (shown in \Fref{fig:d2q9}), which is commonly used for two-dimensional, incompressible flow simulations~\cite{Suc01}. % MG: do we even need this citation?
The lattice is two-dimensional with nine discrete velocities at each node.
There is a stationary particle, there are four discrete velocities of magnitude 1, $(\pm1,0)$, $(0,\pm1)$, and there are four discrete velocities of magnitude $\sqrt{2}$, $(\pm1, \pm1)$.

@@

The discretized version of the Boltzmann equation, or the lattice Boltzmann equation (LBE), is given as: $f_i(\pos + \pvel_i, t + \Delta t) = f_i(\pos, t) + \colop_i(f)$, where, for D2Q9, $i = 0, 1, ..., 8$.

@@

No slip, or zero velocity, which is commonly imposed at walls in a domain, is accomplished by simulating the particle distributions as "bouncing back" at the walls in the opposite direction from which they stream.
For example, for a particle distribution streaming in the direction of a south wall, $f_2 = f_4$, $f_5 = f_7$, and $f_6 = f_8$.
For velocity or pressure boundary conditions, the method proposed by \citet{zou1997pressure} can be used.
The particle distributions that are missing after the streaming step are solved for by assuming a bounceback of the nonequilibrium distribution in the direction normal to the boundary; e.g., for a south inlet or outlet, $f_2 - f_2^{eq} = f_4 - f_4^{eq}$.

@@

\subsection{Collision Operator}

\subsubsection{Bhatnagar-Gross-Krook (BGK)} \label{sec:bgk}

The collision operator, in the case of the continuous BE, attempts to describe the change in particle momentums and trajectories due to pairwise particle collisions (based on their respective momentums and trajectories just prior to collision) \cite{Cer90}.
In LBM, the collision operator causes particle distributions to relax toward a quasiequilibrium. % MG: is this true of all collision operators? or just BGK?
This equilibrium is determined by the macroscopic physical behavior of interest.
In the case of incompressible flow, the quasiequilibrium particle distribution, $f_i^{eq}$, is often given by:

\begin{equation} \label{eq:feq}
f_i^{eq} = w_i \rho \left[1 + \frac{\pvel_i \cdot \mvel}{c_s^2} + \frac{(\pvel_i \cdot \mvel)^2}{2c_s^2} - \frac{\mvel^2}{2c_s^2} \right],
\end{equation}

\noindent where $w_i$ is the weight in the $i^{th}$ direction, and %MG: maybe say something about where this comes from, e.g. feq is a Hermite polynomial expansion of the Maxwell-Boltzmann distribution

\begin{equation} \label{eq:weights}
w_i = \begin{cases}
    \frac{4}{9}, & i = 0 \\
    \frac{1}{9}, & i = 1, 2, 3, 4 \\
    \frac{1}{36}, & i = 5, 6, 7, 8
\end{cases}.
\end{equation}

@@

\begin{equation} \label{eq:bgk}
\Omega_i = -\omega (f_i(\pos, t) - f_i^{eq}),
\end{equation}

\noindent where $\omega$ is the collision frequency \cite{Bha54}.
The collision frequency can be related to macroscopic constitutive properties through Chapman-Enskog multiscale analysis.
For incompressible Newtonian flow, the relaxation time is related to the kinematic viscosity by $\nu = c_s^2(\frac{1}{\omega} - \frac{1}{2})$.

Despite the utility of the BGK collision operator, it does have a few drawbacks.
For example, in low viscosity flows the BGK operator results in an over relaxation of particle distributions toward quasiequilibrium.
It is well known that when large nonequilibrium distributions exist in the LBM approximation that over relaxation can result in nonphysical oscillations that are slow to decay \cite{brownlee2007stability,dellar2003incompressible}.
To illustrate this, consider the example shown in \Fref{fig:sharp-grad}.
If $a >> 0$, then when $f_5(x,y)$ propagates to $f_5(x+1,y+1)$ during the streaming step (where $\pos = [x,y]$, $\pvel_5 = [1,1]$, and $\Delta t = 1$), $f_5(x+1,y+1)$ will be ``far'' from $f^{eq}_5(x+1,y+1)$ for the subsequent collision step.
If $\nu \rightarrow 0$ then $\omega \rightarrow 2$ and \eqref{eq:bgk} results in $f_5(x+1,y+1)$ still being ``far'' from $f^{eq}_5(x+1,y+1)$ but on the ``other side''.
If, as in \Fref{fig:sharp-grad}, the LBM macroscopic variables dictate that $f^{eq}_i$ remain mostly constant (or, at the very least, far from $f_i$) in the direction of propagation for $f_i$, then this leads to oscillations in the particle distribution, $f_i$.
An example of the nonphysical oscillations that would occur for $f_5$ in \Fref{fig:sharp-grad} can be seen in \Fref{fig:overrelaxation}.
Considering the effect oscillations of particle distributions will have on macroscopic variables and, consequently, local quasiequilibriums, positive feedback loops can occur causing the system to ``blow up'' or ``pollute'' the system enough to make the results highly nonphysical and altogether useless \cite{gorban2014enhancement}.

\begin{figure}
    \includegraphics[width=\linewidth]{figs/sharp_gradient}
    \caption{Example of an LBM approximation with a sharp gradient in the macroscopic flow velocity}
    \label{fig:sharp-grad}
\end{figure}

\begin{figure}
    \includegraphics[width=\linewidth]{figs/overrelaxation_tau-0513}
    \caption{Example of oscillations that can occur as a result of overrelaxation (this example is related to the example shown in \Fref{fig:sharp-grad}); the relaxation time, $\omega = 1.95$ and $\nu \approx 0.0043$}
    \label{fig:overrelaxation}
\end{figure}

In contrast to low viscosity flows, the challenge associated with high viscosity flows (using the BGK collision operator) is that particular distributions may never relax as ``far'' toward quasiequilibrium as is physical because as $\nu \rightarrow \infty$, $\omega \rightarrow 0$, resulting in extreme underrelaxation to the point of being negligible. %MG: this needs cleaned up, I think

Concerns with sharp gradients, overrelaxation, and underrelaxation are particularly relevant in non-Newtonian flows because of the nonlinear constitutive relationship between shear stress and strain-rate.
The nonlinear constitutive relationship can lead to sharp gradients, and depending on the form of $\dot{\gamma} \mapsto \mu_{app}$, the apparent viscosity, $\mu(\dot{\gamma}(\pos))$, may result in overrelaxation in certain parts of the domain and extreme underrelaxation in others.

\subsubsection{Multiple-relaxation Time (MRT)}

An alternative to the BGK collision operator is the multiple-relaxation-time (MRT) collision operator.
In the LB-MRT scheme, one constructs a space based on the particle velocity, $\pvel$, moments of $\mathbf{f}$, herein referred to as the ``moment space''.
The collision is then performed in the moment space.
There a few reasons why it is advantageous to perform the collision in the moment space as opposed to the particle distribution space:
\begin{enumerate}
    \item Physical processes in fluids can be approximately described by coupling or interacting among modes; and hydrodynamic modes are directly related to, i.e. linear combinations of, mass and momentum moments \cite{lallemand2000theory}.
    \item For the D2Q9 lattice, there are nine distribution functions, $f_0, f_1, ..., f_8$, but only six variables that affect the intended hydrodynamics on a macroscopic scale, namely: $\rho$, $\mathbf{u}$, and $\mathbf{\Pi}$, where $\mathbf{\Pi}$ is the momentum flux tensor \cite{dellar2003incompressible}.
    Of the nine relaxation rates available, the three that correspond to the extra variables--often referred to as ``ghost variables'', and their associated modes as ``ghost modes''--can be tuned in order to dampen out their associated ghost modes, ensuring these modes do not dominate or cause numerical instabilities at the lattice scale.
\end{enumerate}

The MRT collision operator is given by:

\begin{equation} \label{eq:mrt-colop}
	\colop = - {\transM}^{-1} \relaxM \transM (\mathbf{f} - \mathbf{f}^{eq}),
\end{equation}

\noindent where $\transM$ is a transformation matrix that maps the particle distribution vector, $\mathbf{f}$, and quasiequilibrium distribution vector, $\mathbf{f}^{eq}$, from the particle distribution space into moment space.
The result of mapping the vectors $\mathbf{f}$ and $\mathbf{f}^{eq}$ into moment space will be denoted by $\mathbf{m}$ and $\mathbf{m}^{eq}$, respectively.
The relationships between $\mathbf{m}$, $\transM$ and $\mathbf{f}$ can be written as follows:

\begin{equation} \label{eq:transformation-matrix}
\mathbf{m} = \begin{bmatrix}
\rho \\ e \\ \epsilon \\ j_x \\ q_x \\ j_y \\ q_y \\ p_{xx} \\ p_{xy}
\end{bmatrix} = \begin{bmatrix}
1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
-4 & -1 & -1 & -1 & -1 & 2 & 2 & 2 & 2 \\
4 & -2 & -2 & -2 & -2 & 1 & 1 & 1 & 1 \\
0 & 1 & 0 & -1 & 0 & 1 & -1 & -1 & 1 \\
0 & -2 & 0 & 2 & 0 & 1 & -1 & -1 & 1 \\
0 & 0 & 1 & 0 & -1 & 1 & 1 & -1 & -1 \\
0 & 0 & -2 & 0 & 2 & 1 & 1 & -1 & -1 \\
0 & 1 & -1 & 1 & -1 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 1 & -1 & 1 & -1
\end{bmatrix} \begin{bmatrix}
f_0 \\
f_1 \\
f_2 \\
f_3 \\
f_4 \\
f_5 \\
f_6 \\
f_7 \\
f_8
\end{bmatrix} = \transM \mathbf{f}.
\end{equation}

\noindent where $\epsilon$ is related to the square of the energy $e$; $q_x$ and $q_y$ correspond to the energy fluxes in the x and y directions; and $p_{xx}$ and $p_{xy}$ correspond to the diagonal and off-diagonal component of the viscous stress tensor~\cite{lallemand2000theory}. % MG: verify this is correct
The relaxation matrix, $\relaxM$, is a diagonal matrix where each of the elements on the diagonal, $s_i, i = 0, 1, ..., 8$, correspond to the relaxation rate of its associated hydrodynamic mode.
In the case when $s_0 = s_1 = ... = s_8 = \omega$ ($\omega$ is the collision frequency in the BGK sense), the MRT collision operator is equivalent to the BGK collision operator.
The relaxation parameters $s_0$, $s_3$, and $s_5$ are all set to zero as mass and momentum should be conserved.
The relaxation parameters $s_1$ and $s_7 = s_8$ are related to the bulk and shear viscosities, respectively.
The relationship for the shear viscosity is given by,
\begin{equation} \label{eq:mrt-s7}
\nu = c_s^2 \Delta t (\frac{1}{s_7} - \frac{1}{2}).
\end{equation}
The remaining relaxation parameters, $s_2$, $s_4$, and $s_6$, are tuned in order to dampen out and separate the ghost modes from the modes affecting hydrodynamic transport.
It is common practice, and \citet{lallemand2000theory} recommends, that these three relaxation parameters be set to slightly larger than one.

The MRT collision operator has a greater numerical stability than its BGK counterpart~\cite{lallemand2000theory,dellar2003incompressible,d2002multiple}, and because of the challenges associated with simulating non-Newtonian flow, the MRT collision operator has become popular for simulating non-Newtonian flows~\cite{chen2014simulations,fallah2012multiple,tang2011bingham,vikhansky2008lattice,chai2011multiple,grasinger2015simulation}.
The main drawback of the MRT collision operator is its computational expense.
Why MRT is more computationally expense is clear when one considers that \eqref{eq:mrt-colop} requires multiple matrix multiplications and \eqref{eq:bgk} requires none.
It has been reported that MRT is approximately 15\% slower than BGK~\cite{d2002multiple}, but this was in the context of Newtonian flow.
As will be shown later, for certain LBM implementations and non-Newtonian flows the increase in computational expense can be much greater.

\subsection{Stability Enhancement through Artificial Dissipation}

To reduce nonequilibrium fluctuations in LBM, one can introduce artificial dissipation.
The idea of artificial dissipation is to increase numerical stability, while sacrificing some physical accuracy.
A model that has more physical justification but produces unstable and nonsensical results is much less useful than a model with some minor artificial features yet produces more stable results.
%Lyapanov stability is the idea that solutions at equilibrium stay there.
A practical goal then would be to use only the necessary amount of artificial dissipation in order to ensure a stable solution.
From this goal two questions naturally arise: ``under what criteria does one decide that artificial dissipation necessary?'' and ``how much artificial dissipation does one introduce when it is necessary?''.
